Geospatial - Assignment 2

### 2025-02-03 hw2/hw2_MoritzPeistNoemiLucchiSimonVellin.Rmd
Added Simon's code - TODO: Add documentation in Part @Simon

# Spatial packages
library(sf)
library(spData)
library(tidyverse)
library(units)
library(lwgeom)
```

@@

>>>>

# Spatial packages
library(sf)
library(spData)
library(tidyverse)
library(units)
library(lwgeom)
library(gridExtra)
library(readxl)
```

@@

In this exercise, we use the [Porteous dataset]() for plotting.

## 2.1. Map of markets in Africa

```{r}
# Clean enviornment first
rm(list = ls())

# Data Loading (with correct CRS and long/lat conversions)
sf.world <- st_read("data/ne_10m_admin_0_countries", crs = "EPSG:4326")
sf.roads_africa <- st_read("data/ne_10m_roads", crs = "EPSG:4326")
sf.markets <- st_as_sf(
  read_excel("data/porteous_data/1.-Price--Production--and-Population-Data/MktCoords.xlsx"),
  coords = c("longitude", "latitude"),
  crs = "EPSG:4326"
)
```

```{r}
# Plot
ggplot() +
  geom_sf(
    data = sf.world %>% filter(REGION_UN == "Africa"),
    fill = "antiquewhite", color = "gray70"
  ) +
  geom_sf(data = sf.markets, color = "red", size = 0.7, shape = 21, fill = "blue") +
  theme_minimal() +
  labs(
    title = "Markets in Africa",
    caption = "Source: Replication Data provided"
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5),
    plot.caption = element_text(size = 10, hjust = 0.5)
  )
```

## 2.2. Minimum distances

```{r}
# Load data
sf.airports <- st_read("data/ne_10m_airports", crs = "EPSG:4326")
sf.coast <- st_read("data/ne_10m_coastline", crs = "EPSG:4326")
```

```{r}
# We isolate the African airports from our global data sets to stay efficient.
sf.africa <- sf.world %>%
  filter(REGION_UN == "Africa") %>%
  filter(!st_is_valid(.)) %>%
  st_make_valid()

# Filter airports in Africa
sf.airports_africa <- sf.airports %>% st_filter(sf.africa)

# Data Wrangling - st_distance() returns a matrix with one row per market and one column per coast segment.

# (i) Distance to Nearest Coast
dist_to_coast <- st_distance(sf.markets, sf.coast)
# For each market (row) we take the minimum distance:
min_coast_dist <- apply(dist_to_coast, 1, min)

# (ii) Distance to Nearest Road
dist_to_roads <- st_distance(sf.markets, sf.roads_africa)
min_road_dist <- apply(dist_to_roads, 1, min)

# (iii) Distance to Nearest Airport
dist_to_airports <- st_distance(sf.markets, sf.airports_africa)
min_airport_dist <- apply(dist_to_airports, 1, min)

# Add the computed distances to the markets data
sf.markets <- sf.markets %>%
  mutate(
    min_coast_dist   = as.numeric(min_coast_dist),
    min_road_dist    = as.numeric(min_road_dist),
    min_airport_dist = as.numeric(min_airport_dist)
  )

# Summary
print(summary(sf.markets))
```

## 2.3. Scatter plots

```{r}
prices <- read_excel(
  "data/porteous_data/4.-Outer-Loop-for-Cost-Parameter-Estimation/PriceMaster4GAMS.xlsx"
)
```

```{r}
# Calculate average price
prices_slice <- prices[, 5:ncol(prices)]
prices <- replace(prices, prices == 0, NA) %>%
  mutate(av_price = rowMeans(prices_slice, na.rm = TRUE)) %>%
  select(mktcode, country, market, crop, av_price)
sf.marketprices <- sf.markets %>% left_join(prices, by = "mktcode")
```

```{r}
# (i) Average Price vs. Minimum Distance to Coast
p_coast <- ggplot(sf.marketprices, aes(x = min_coast_dist, y = av_price, color = crop)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) + # Add linear trend lines
  labs(
    title = "Average Price vs. Distance to Coast",
    x = "Minimum Distance to Coast (meters)",
    y = "Average Price",
    color = "Crop Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

# (ii) Average Price vs. Minimum Distance to Road
p_road <- ggplot(sf.marketprices, aes(x = min_road_dist, y = av_price, color = crop)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) + # Add linear trend lines
  labs(
    title = "Average Price vs. Distance to Road",
    x = "Minimum Distance to Road (meters)",
    y = "Average Price",
    color = "Crop Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

# (iii) Average Price vs. Minimum Distance to Airport
p_airport <- ggplot(sf.marketprices, aes(x = min_airport_dist, y = av_price, color = crop)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) + # Add linear trend lines
  labs(
    title = "Average Price vs. Distance to Airport",
    x = "Minimum Distance to Airport (meters)",
    y = "Average Price",
    color = "Crop Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

# Display the plots in a grid
grid.arrange(p_coast, p_road, p_airport, nrow = 2, ncol = 2)
```
